# Reproducible Research: Peer Assessment 1

This is a complete report for the project.
All displayed code is written assuming that a project dir is a working dir.

## Loading and preprocessing the data

First we read an unzipped dataset choosing appropriate formats for variables.
```{r readzipfile}
data<-read.csv(unzip("activity.zip"),colClasses=c("numeric","Date","factor"))
```

## What is mean total number of steps taken per day?

This chunk of code plots histograms of steps taken each day using lattice package.
```{r histogram}
library(lattice)
histogram(~steps|as.factor(date),data,par.strip.text=list(cex=0.7))
```

And this piece computes and returns a table with daily mean and median values.
```{r meanmedian}
steps.mm<-as.data.frame(sapply(split(data$steps,as.factor(data$date)),mean,na.rm=T))
names(steps.mm)<-"Mean"
steps.mm$Median<-sapply(split(data$steps,as.factor(data$date)),median,na.rm=T)
print(steps.mm)
```

## What is the average daily activity pattern?

First we compile a data frame with average number of steps per interval.
```{r stepsperinterval}
steps.int<-as.data.frame(sapply(split(data$steps,data$interval),mean,na.rm=T))
steps.int$interval<-as.numeric(rownames(steps.int))
names(steps.int)<-c("steps","interval")
steps.int<-steps.int[order(steps.int$interval),]
```

Using the data frame constructed above one can build a graph with average number of steps taken in each interval.
```{r stepsperinterval_chart}
plot(x=steps.int$interval,y=steps.int$steps,type="l",xlab="Interval",ylab="Number of steps",lwd=2)
```

The following chunck computes and properly formats a 5-min interval with a max average number of steps.
```{r maxstepsperinterval}
m<-as.character(steps.int[(order(steps.int$steps,decreasing=T)),][1,2])
max<-paste0(substr(m,1,1),":",substr(m,2,3))
```

The 5-min interval with a maximum average number of steps is an interval starting at `r max`.

## Imputing missing values

This piece calculates the total number of missing entries.
```{r totalnas}
tot.na<-sum(is.na(data$steps))
```

There're total `r tot.na` NAs in the data set.

The following chunk replaces NAs in the original dataset with average values per corresponding time interval using the 'plyr' package. New 'clean' dataset is saved.
```{r replacenas}
library(plyr)
data.nonas<-join(data,steps.int,"interval")
for(i in seq_along(data$steps)){
      if(is.na(data.nonas[i,1])==TRUE){
            data.nonas[i,1]<-data.nonas[i,4]
      }
}
data.nonas<-data.nonas[,1:3]
```

Using a 'clean' dataset created above we calculate daily sums of steps taken.
```{r summeanmedian_nonas}
steps.sum.nonas<-as.data.frame(sapply(split(data.nonas$steps,as.factor(data.nonas$date)),sum,na.rm=T))
names(steps.sum.nonas)<-"Sum"
```

Here's a histogram of a total number of daily steps.
```{r histsum_nonas}
hist(steps.sum.nonas$Sum,xlab="",main="Total daily number of steps histogram")
```

Mean total daily steps taken (rounded) are `r as.integer(mean(steps.sum.nonas$Sum))`, median is `r as.integer(median(steps.sum.nonas$Sum))`. Both values are higher compared to ones computed ignoring NAs in the dataset. This is due to chosen NAs replacement strategy (average values per corresponding interval).

## Are there differences in activity patterns between weekdays and weekends?

Let's first create a new factor variable with 'weekday' and 'weekend' levels.
```{r weekdays}
data.nonas$weekday<-weekdays(data.nonas$date)
for(i in seq_along(data.nonas$weekday)){
      if(data.nonas$weekday[i]=="Saturday"){
            data.nonas$weekday[i]<-"weekend"
      }else if(data.nonas$weekday[i]=="Sunday"){
            data.nonas$weekday[i]<-"weekend"
      }else{
            data.nonas$weekday[i]<-"weekday"
      }
}
data.nonas$weekday<-as.factor(data.nonas$weekday)
```

Now we create a new data frame with average number of steps per interval splitting out weekdays and weekends.
```{r weekdays_split}
l<-split(data.nonas,data.nonas$weekday)
steps.int.nonas<-as.data.frame(sapply(l,function(x){sapply(split(x,x[,3]),function(y){mean(y[,1])})}))
steps.int.nonas$interval<-as.numeric(rownames(steps.int.nonas))
steps.int.nonas<-steps.int.nonas[order(steps.int.nonas$interval),]
```

Finally we plot average number of steps per time interval for weekdays and weekends.
```{r weekdays_plot}
par(mfrow=c(1,2))
plot(x=steps.int.nonas$interval,y=steps.int.nonas$weekday,type="l",xlab="Interval",ylab="Number of steps",main="Weekdays",ylim=c(0,250),cex.axis=0.8)
plot(x=steps.int.nonas$interval,y=steps.int.nonas$weekend,type="l",xlab="Interval",ylab="Number of steps",main="Weekends",ylim=c(0,250),cex.axis=0.8)
```
